\\textsc{Mathematical Description}

### 2023-06-07 100-math-description/GR-sslcm-math-description.Rmd
Edits to math document to reflect recent model changes
See check list in #183

---
title: "\\textsc{Mathematical Description}"
subtitle: "\\textsc{State-Space Life Cycle Model for Grande Ronde Spring Chinook Salmon}"
output: 
  bookdown::pdf_document2:
    keep_tex: false
    toc: true
    toc_depth: 3
    number_sections: true
    citation_package: natbib
  # bookdown::word_document2
fontfamily: mathpazo
geometry: [top=1in,bottom=1in,left=0.25in,right=1.75in]
linkcolor: cyan
urlcolor: cyan
citecolor: black
bibliography: cites.bib
link-citations: true
biblio-style: apalike
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{microtype}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \definecolor{my-blue}{HTML}{E6F0FF}
  - \usepackage[textwidth=1.5in]{todonotes}
  - \setuptodonotes{backgroundcolor=my-blue,textcolor=black,linecolor=ProcessBlue,bordercolor=my-blue,size=footnotesize}
  - \usepackage[labelfont={bf,sc},labelsep=period]{caption}
  - \captionsetup{width=\textwidth}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \cfoot{\thepage}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \lfoot{State-Space Life Cycle Model\\Mathematical Description}
  - \rfoot{Grande Ronde Spring Chinook Salmon}
  - \usepackage{titlesec}
  - \titleformat{\section}{\bfseries\LARGE\scshape}{\thesection}{1em}{}
  - \titleformat{\subsection}{\bfseries\large}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\itshape}{\thesubsubsection}{1em}{}
  - \renewcommand{\contentsname}{Table of Contents}
  - \renewcommand{\eqref}[1]{\ref{#1}}
  - \renewcommand*{\log}{\mathrm{log}\hspace{-0.20em}}
  - \renewcommand*{\max}{\mathrm{max}\hspace{-0.20em}}
  - \newcommand{\logit}{\mathrm{logit}\hspace{-0.20em}}
  - \newcommand{\ilogit}{\mathrm{logit}^{-1}\hspace{-0.20em}}
  - \newcommand{\MVNdist}{\mathcal{MVN}\hspace{-0.20em}}
  - \newcommand{\Ndist}{\mathcal{N}\hspace{-0.20em}}
  - \newcommand{\Bdist}{\mathcal{B}\hspace{-0.20em}}
  - \newcommand{\Mdist}{\mathcal{M}\hspace{-0.20em}}
---

@@

where $\Sigma_{\phi^{A \rightarrow B}}$ is an $n_j \times n_j$ matrix with the structure:

\begin{equation}
\left[
\begin{array}{cccc}
\left(\sigma_{\phi^{A \rightarrow B}_{j=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{j=2}}\right)^2 & \dots & \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \left(\sigma_{\phi^{A \rightarrow B}_{j=n_j}}\right)^2
\end{array}
\right]
(\#eq:generic-Sigma4by4)
\end{equation}

That is, all populations ($j$) have unique variance terms $(\sigma_{\phi^{A \rightarrow B}_j})^2$ but we assume the correlation component of the covariance calculation ($\rho_{\phi^{A \rightarrow B}}$) is identical for all population pairs.
In cases where a quantity varies annually and is different by origin but not by population, $\Sigma$ would have dimensions $n_o \times n_o$:

@@

Although the stochastic process requires the $\Sigma$ term, is is the component $\sigma_j$ (or $\sigma_o$) and $\rho$ terms that were assigned priors ($\mathcal{U}(0,5)$ and $\mathcal{U}(-1,1)$, respectively, Table \@ref(tab:param-table)) and the $\Sigma$ matrices were constructed using the structures defined in eqs. \@ref(eq:generic-Sigma4by4) and \@ref(eq:generic-Sigma2by2).

@@

\begin{equation}
E_{y,j} = \sum_o^{n_o} \sum_k^{n_k} S^{a}_{y,k,o,j} \cdot \Omega_{k} \cdot f_{k}
(\#eq:get-E)
\end{equation}

We assumed that spawners age-4 or age-5 were 50% female and age-3 spawners were all male (i.e., $\Omega_{k=1} = 0;\;\Omega_{k \in [2,3]} = 0.5$; approximately equal to values estimated from carcass data) and estimated fecundity ($f_k$) from samples taken from Grande Ronde-origin hatchery broodstock.
We supplied both quantities to the model assuming no error or inter-annual variability.

@@

We modeled realized (i.e., with process noise) egg-to-parr survival as a multivariate logit-normal random variable around the expected value ($\dot{\phi}^{E \rightarrow P^{b}}_{y,j}$) from eq. \@ref(eq:get-phi-E-Pb-dot).
Rather than assume that inter-annual variability in egg-to-parr survival beyond that explained by density-dependence is completely random across years, we allowed the model to account for serial autocorrelation in egg survival anomalies via a lag-1 autoregressive process [AR(1); coefficient denoted $\kappa^{E \rightarrow P^{b}}_{j}$], wherein the white noise portion of the anomaly has covariance matrix $\Sigma_{\phi^{E \rightarrow P^{b}}}$ (constructed of $\sigma_{\phi^{E \rightarrow P^{b}}_j}$ and $\rho_{\phi^{E \rightarrow P^{b}}}$ following eq. \@ref(eq:generic-Sigma4by4)):

@@

and the realized parr mean length as being multivariate log-normally distributed around $\dot{L}^{P^{b}}_{y,j}$ with covariance matrix $\Sigma_{L^{P^{b}}}$ (constructed of $\sigma_{L_{j}^{P^{b}}}$ and $\rho_{L^{P^{b}}}$ following eq. \@ref(eq:generic-Sigma4by4)):

@@

where $P^{a}_{y,i,j}$ is migratory strategy-specific parr abundance and $\pi_{y,i,j}$ is the proportion that take on each migratory strategy.
We assumed the expected value of this proportion for fall migrants ($\dot{\pi}_{i = \mathrm{fall},j}$) was time-constant and modeled the realized values as having a multivariate logit-normal distribution with covariance matrix $\Sigma_{\pi_{i = \mathrm{fall}}}$ (constructed of $\sigma_{\pi_{i=\mathrm{fall},j}}$ and $\rho_{\pi_{i=\mathrm{fall}}}$ following eq. \@ref(eq:generic-Sigma4by4)):

@@

\begin{equation}
\logit\left(\dot{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,j}\right) = \gamma_{0,i,j} + \gamma_{1,i,j} \cdot L^{*P^{b}}_{y,j}
(\#eq:get-phi-Pa-Mb-dot)
\end{equation}

where $L^{*P^{b}}_{y,j}$ is parr mean length scaled and centered by subtracting the historical average observed parr mean length for population $j$ from $L^{P^{b}}_{y,j}$ and dividing by the historical observed standard deviation.
Then, for each migratory strategy separately, overwinter survival was a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{P^{a} \rightarrow M^{b}}_i}$ (each matrix constructed of $\sigma_{\phi^{P^{a} \rightarrow M^{b}}_{i,j}}$ and $\rho_{\phi^{P^{a} \rightarrow M^{b}}_{i}}$ following eq. \@ref(eq:generic-Sigma4by4)):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\phi}}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right), \Sigma_{\phi^{P^{a} \rightarrow M^{b}}_i}\right]
(\#eq:phi-Pa-Mb-random)
\end{equation}

@@

@@

and obtained the realized multiplicative change in mean length as being multivariate log-normally distributed around $\dot{\Delta}^{L^{P^{b}} \rightarrow L^{M^{b}}}_{y,j}$ with covariance matrix $\Sigma_{\Delta^{L^{P^{b}} \rightarrow L^{M^{b}}}}$ (constructed of $\sigma_{\Delta_{j}^{L^{P^{b}} \rightarrow L^{M^{b}}}}$ and $\rho_{\Delta^{L^{P^{b}} \rightarrow L^{M^{b}}}}$ following eq. \@ref(eq:generic-Sigma4by4)):

@@

where $L^{*M^{b}}_{y,j}$ has been scaled and centered by subtracting the historical average observed smolt mean length for population $j$ from $L^{M^{b}}_{y,j}$ and dividing by the historical observed standard deviation (just as for $L^{*P^{b}}_{y,j}$ for standardized parr mean length).
For hatchery-origin smolt, we assumed this expected out-of-basin survival term was time-constant (i.e., all $\dot{\phi}^{M^{b} \rightarrow M^{a}}_{y,o=\mathrm{HOR},j}$ were equal within a population).
For each origin separately, we modeled realized out-of-basin migration survival as a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{M^{b} \rightarrow M^{a}}_o}$ (each matrix constructed of $\sigma_{\phi^{M^{b} \rightarrow M^{a}}_{o,j}}$ and $\rho_{\phi^{M^{b} \rightarrow M^{a}}_o}$ following eq. \@ref(eq:generic-Sigma4by4)):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{b} \rightarrow M^{a}}_{y,o,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{b} \rightarrow M^{a}}_{y,o,1:n_j}\right), \Sigma_{\phi^{M^{b} \rightarrow M^{a}}_o}\right]
(\#eq:phi-Mb-Ma-random)
\end{equation}

@@

We modeled survival during the downstream migration through the hydrosystem on the Snake and Columbia rivers as a single survival rate, specific to each year and origin but shared among populations ($\phi^{M^{a} \rightarrow O^{0}}_{y,o}$).
We assumed these survival rates were multivariate logit-normal random variables around a time-constant expected value ($\dot{\phi}^{M^{a} \rightarrow O^{0}}_{o}$) with covariance matrix $\Sigma_{\phi^{M^{a} \rightarrow O^{0}}}$ (constructed of $\sigma_{\phi^{M^{a} \rightarrow O^{0}}_{o}}$ and $\rho_{\phi^{M^{a} \rightarrow O^{0}}}$ following eq. \@ref(eq:generic-Sigma2by2)):

@@

The observation model components for the freshwater juvenile life stage were fitted to three primary data types: (_i_) abundance of juveniles passing rotary smolt traps (located in each tributary, \todo{Include reference to document describing methods of juvenile passage estimation}) in either the fall or spring, (_ii_) survival of juveniles from several PIT tagging events to LGR \todo{Include reference to document describing methods of juvenile survival estimation based on PIT tags}, and (iii) the mean length of sampled individuals during two of the PIT tagging events \todo{Include reference to document describing methods of juvenile mean length estimation}.
For all types, the information supplied to the model were externally compiled estimates (e.g., survival from Cormack-Jolly-Seber models) and the estimated standard error was supplied to the model as a measure of observation error variability.
A more fully integrated model would involve constructing the likelihood based on the frequency of PIT tag detection histories, however we thought unnecessary given the pre-compiled nature of currently available estimates and their uncertainty  \todo{Consider citing Staton et al. (2017) integrated vs. sequential SSSRA paper here}.

@@

We assumed first year natural-origin^[All $o=\mathrm{NOR}$ in eq. \@ref(eq:phi-O0-O1-random), omitted for brevity.] ocean survival was a multivariate logit-normal random variable around a time-constant expected value ($\dot{\phi}^{O^{0} \rightarrow O^{1}}_{o=\mathrm{NOR},j}$) with covariance matrix $\Sigma_{\phi^{O^{0} \rightarrow O^{1}}}$ (constructed of $\sigma_{\phi^{O^{0} \rightarrow O^{1}}_{j}}$ and $\rho_{\phi^{O^{0} \rightarrow O^{1}}}$ following eq. \@ref(eq:generic-Sigma4by4)), but included a lag-1 autoregressive process as a means to account for ocean conditions that may affect survival in a temporally non-independent fashion\todo{Include reference to papers that talk about ocean survival varying cyclically; e.g., Mantua et al. paper on PDO}:

@@

We treated realized maturation rates (i.e., the proportion of fish from a given brood year alive and in the ocean at the beginning of a year that make the return migration that year) as multivariate logit-normal random variables around time-constant expected values ($\dot{\psi}^{O^{w}}_{o,j}$; $w$ is the number of winters spent in the ocean) with covariance matrix $\Sigma_{\psi^{O^{w}}_o}$ (one matrix per origin, each constructed of $\sigma_{\psi^{O^{w}}_{o,j}}$ and $\rho_{\psi^{O^{w}}_{o}}$ following eq. \@ref(eq:generic-Sigma4by4)):

\begin{equation}
\begin{aligned}
\logit\left(\boldsymbol{\psi}^{O^{1}}_{y,o,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\psi}}^{O^{1}}_{o,1:n_j}\right), \Sigma_{\psi^{O^{1}}_o}\right] \\
\logit\left(\boldsymbol{\psi}^{O^{2}}_{y,o,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\psi}}^{O^{2}}_{o,1:n_j}\right), \Sigma_{\psi^{O^{2}}_o}\right]
\end{aligned}
(\#eq:psi-random)
\end{equation}

@@

The abundances of ocean juveniles ($O^{1}_{y,o,j}$, $O^{2}_{y,o,j}$, and $O^{3}_{y,o,j}$) are organized by brood year, however they will return in different years to spawn (and be observed) because they are of different ages.
We placed returning mature fish into the correct year and age of return to the mouth of the Columbia River before the upstream migration ($R^{b}_{y,k,o,j}$) based on the following rule.
Fish spawned in year $y$ will return as the $k^{\mathrm{th}}$ possible age of maturity in year $y + k_{\mathrm{min}} + k - 1$ (age-3 is the first possible total age of maturity, thus $k_{\mathrm{min}}$ = 3).
As an example, fish spawned in brood year $y$ = 2000 will return in 2003 as age-3 ($k$ = 1), in 2004 as age-4 ($k$ = 2), or in 2005 as age-5 ($k$ = 3) -- see Table \@ref(tab:year-table) for a visual of this example.
Thus, the abundance of fish spawned in brood year $y$ returning as age-3, age-4, and age-5, respectively, was:

\begin{equation}
\begin{aligned}
R^{b}_{y+k_{\mathrm{min}} + 1 - 1,1,o,j} &= O^{1}_{y,o,j} \cdot \psi^{O^{1}}_{y,o,j} \\
R^{b}_{y+k_{\mathrm{min}} + 2 - 1,2,o,j} &= O^{2}_{y,o,j} \cdot \psi^{O^{2}}_{y,o,j} \\
R^{b}_{y+k_{\mathrm{min}} + 3 - 1,3,o,j} &= O^{3}_{y,o,j} \cdot \psi^{O^{3}}_{y,o,j} \\
\end{aligned}
(\#eq:get-Rbs)
\end{equation}

@@

Prior to reaching BON, we subject the returning adults to predation mortality from sea lions (survival denoted by $\phi^{SL}_{y,j}$) and harvest mortality from fisheries (harvest rate on fish surviving sea lion predation denoted by $U_{y,k,o}$).\todo{Insert information about where harvest rates and sea lion predation estimates came from.}
We compiled the best available estimates of these quantities, but did not fit to them explicitly as part of the observation model (as we do with many other data sources) -- instead we passed them to the model as non-stochastic quantities that were assumed to be known without error (used in eq. \@ref(eq:get-Ra), below).

@@

Once fish reach BON, however, their survival to LGR has been monitored (since 2000; $y=10$) via PIT tags and so we did fit to those data and estimate the survival rate between BON and LGR as multivariate logit-normal random variables around expected values ($\dot{\phi}^{R^{b} \rightarrow R^{a}}_{o}$) with covariance matrix $\Sigma_{\phi^{R^{b} \rightarrow R^{a}}}$ (constructed of $\sigma_{\phi^{R^{b} \rightarrow R^{a}}_o}$ and $\rho_{\phi^{R^{b} \rightarrow R^{a}}}$ following eq. \@ref(eq:generic-Sigma2by2)).
In years without these PIT tag counts, however, the survival from BON to LGR is confounded with ocean survival, and so we assumed BON to LGR survival was equal to the expected value in these years for parameter identifiability purposes:

@@

\begin{equation}
R^{a}_{y,k,o,j} = R^{b}_{y,k,o,j} \cdot \phi^{SL}_{y,j} \cdot (1 - U_{y,k,o}) \cdot \phi^{R^{b} \rightarrow R^{a}}_{y,o} + \left(G_{y,o,j} \cdot p^{G}_{k,j}\right)
(\#eq:get-Ra)
\end{equation}

\todo{Include note that no mortality is assumed between LGR and the natal tributary.}

@@

Upon arrival to the natal tributary, populations with a weir and hatchery program (all populations except the Minam River, indexed by $j = \mathrm{MIN}$) have some number of adults removed each year for broodstock and as a means to minimize natural spawning of hatchery-origin fish.
We supplied these weir removals (denoted by $B_{y,k,o,j}$) to the model as known without error and set all $B_{y,k,o,j=\mathrm{MIN}} = 0$.
Thus, the abundance of adults that would have a chance to spawn each year was (the maximum constraint was used to ensure all $S^{b}_{y,k,o,j} > 0$):

>>>>

---
title: "\\textsc{Mathematical Description}"
subtitle: "\\textsc{State-Space Life Cycle Model for Grande Ronde Spring Chinook Salmon}"
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: true
    toc_depth: 3
    number_sections: true
    citation_package: natbib
  # bookdown::word_document2
fontfamily: mathpazo
geometry: [top=1in,bottom=1in,left=0.25in,right=1.75in]
linkcolor: cyan
urlcolor: cyan
citecolor: black
bibliography: cites.bib
link-citations: true
biblio-style: apalike
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{microtype}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \definecolor{my-blue}{HTML}{E6F0FF}
  - \usepackage[textwidth=1.5in]{todonotes}
  - \setuptodonotes{backgroundcolor=my-blue,textcolor=black,linecolor=ProcessBlue,bordercolor=my-blue,size=footnotesize}
  - \usepackage[labelfont={bf,sc},labelsep=period]{caption}
  - \captionsetup{width=\textwidth}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \cfoot{\thepage}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \lfoot{State-Space Life Cycle Model\\Mathematical Description}
  - \rfoot{Grande Ronde Spring Chinook Salmon}
  - \usepackage{titlesec}
  - \titleformat{\section}{\bfseries\LARGE\scshape}{\thesection}{1em}{}
  - \titleformat{\subsection}{\bfseries\large}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\itshape}{\thesubsubsection}{1em}{}
  - \renewcommand{\contentsname}{Table of Contents}
  - \renewcommand{\eqref}[1]{\ref{#1}}
  - \renewcommand*{\log}{\mathrm{log}\hspace{-0.20em}}
  - \renewcommand*{\max}{\mathrm{max}\hspace{-0.20em}}
  - \newcommand{\logit}{\mathrm{logit}\hspace{-0.20em}}
  - \newcommand{\ilogit}{\mathrm{logit}^{-1}\hspace{-0.20em}}
  - \newcommand{\MVNdist}{\mathcal{MVN}\hspace{-0.20em}}
  - \newcommand{\Ndist}{\mathcal{N}\hspace{-0.20em}}
  - \newcommand{\Bdist}{\mathcal{B}\hspace{-0.20em}}
  - \newcommand{\Mdist}{\mathcal{M}\hspace{-0.20em}}
---

@@

where $\Sigma_{\phi^{A \rightarrow B}}$ is an $n_j \times n_j$ matrix constructed of population-specific variance terms $\left(\sigma_{\phi^{A \rightarrow B}_j}\right)^2$ and correlation terms specific to each pair of populations $j$ and $j^\prime$ $\rho_{\phi^{A \rightarrow B}_{j,j^{\prime}}}$:

\begin{equation}
\left[
\begin{array}{cccc}
\left(\sigma_{\phi^{A \rightarrow B}_{j=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}_{j=1,j^{\prime}=2}} & \dots & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}_{j=1,j^{\prime}=n_j}} \\
\sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}_{j=2,j^{\prime}=1}} & \left(\sigma_{\phi^{A \rightarrow B}_{j=2}}\right)^2 & \dots & \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}_{j=2,j^{\prime}=n_j}} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}_{j=n_j,j^{\prime}=1}} & \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}_{j=n_j,j^{\prime}=2}} & \dots & \left(\sigma_{\phi^{A \rightarrow B}_{j=n_j}}\right)^2
\end{array}
\right]
(\#eq:generic-Sigma4by4)
\end{equation}

In cases where a quantity varies annually and is different by origin but not by population, $\Sigma$ would have dimensions $n_o \times n_o$:

@@

We estimated these covariance matrices by assigning their inverse ($\Sigma^{-1}$) a scaled Wishart prior distribution (Table \@ref(tab:param-table)).
Although the stochastic process requires the $\Sigma$ term, is is the component $\sigma$ and $\rho$ terms that we were interested in for inference.

@@

\begin{equation}
E_{y,j} = \sum_o^{n_o} \sum_k^{n_k} S^{a}_{y,k,o,j} \cdot \Omega_{k} \cdot f_{y,k,j}
(\#eq:get-E)
\end{equation}

We assumed that spawners age-4 or age-5 were 50% female and age-3 spawners were all male (i.e., $\Omega_{k=1} = 0;\;\Omega_{k \in [2,3]} = 0.5$; approximately equal to values estimated from carcass data) and estimated fecundity ($f_{y,k,j}$) from length-fecundity relationships based on Grande Ronde-origin hatchery broodstock.
We supplied both quantities to the model assuming no error or inter-annual variability.

@@

We modeled realized (i.e., with process noise) egg-to-parr survival as a multivariate logit-normal random variable around the expected value ($\dot{\phi}^{E \rightarrow P^{b}}_{y,j}$) from eq. \@ref(eq:get-phi-E-Pb-dot).
Rather than assume that inter-annual variability in egg-to-parr survival beyond that explained by density-dependence is completely random across years, we allowed the model to account for serial autocorrelation in egg survival anomalies via a lag-1 autoregressive process [AR(1); coefficient denoted $\kappa^{E \rightarrow P^{b}}_{j}$], wherein the white noise portion of the anomaly has covariance matrix $\Sigma_{\phi^{E \rightarrow P^{b}}}$:

@@

and the realized parr mean length as being multivariate log-normally distributed around $\dot{L}^{P^{b}}_{y,j}$ with covariance matrix $\Sigma_{L^{P^{b}}}$:

@@

where $P^{a}_{y,i,j}$ is migratory strategy-specific parr abundance and $\pi_{y,i,j}$ is the proportion that take on each migratory strategy.
We assumed the expected value of this proportion for fall migrants ($\dot{\pi}_{i = \mathrm{fall},j}$) was time-constant and modeled the realized values as having a multivariate logit-normal distribution with covariance matrix $\Sigma_{\pi_{i = \mathrm{fall}}}$:

@@

\begin{equation}
\logit\left(\dot{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,j}\right) = \gamma_{0,i,j} + \gamma_{1,j} \cdot L^{*P^{b}}_{y,j}
(\#eq:get-phi-Pa-Mb-dot)
\end{equation}

where $L^{*P^{b}}_{y,j}$ is parr mean length scaled and centered by subtracting the historical average observed parr mean length for population $j$ from $L^{P^{b}}_{y,j}$ and dividing by the historical observed standard deviation in mean length across years.
Then, for each migratory strategy separately, overwinter survival was a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{P^{a} \rightarrow M^{b}}}$ (assumed common among migratory strategies):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\phi}}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right), \Sigma_{\phi^{P^{a} \rightarrow M^{b}}}\right]
(\#eq:phi-Pa-Mb-random)
\end{equation}

@@

$\rho_{\pi_{i=\mathrm{fall},j,j^{\prime}}}$

@@

and obtained the realized multiplicative change in mean length as being multivariate log-normally distributed around $\dot{\Delta}^{L^{P^{b}} \rightarrow L^{M^{b}}}_{y,j}$ with covariance matrix $\Sigma_{\Delta^{L^{P^{b}} \rightarrow L^{M^{b}}}}$:

@@

where $L^{*M^{b}}_{y,j}$ has been scaled and centered by subtracting the historical average observed smolt mean length for population $j$ from $L^{M^{b}}_{y,j}$ and dividing by the historical observed standard deviation (just as for $L^{*P^{b}}_{y,j}$ for standardized parr mean length).
For hatchery-origin smolt, we assumed this expected out-of-basin survival term was time-constant (i.e., all $\dot{\phi}^{M^{b} \rightarrow M^{a}}_{y,o=\mathrm{HOR},j}$ were equal within a population).
For each origin separately, we modeled realized out-of-basin migration survival as a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{M^{b} \rightarrow M^{a}}}$ (assumed common across origin types):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{b} \rightarrow M^{a}}_{y,o,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{b} \rightarrow M^{a}}_{y,o,1:n_j}\right), \Sigma_{\phi^{M^{b} \rightarrow M^{a}}}\right]
(\#eq:phi-Mb-Ma-random)
\end{equation}

@@

We modeled survival during the downstream migration through the hydrosystem on the Snake and Columbia rivers as a single survival rate, specific to each year and origin but shared among populations ($\phi^{M^{a} \rightarrow O^{0}}_{y,o}$).
We assumed these survival rates were multivariate logit-normal random variables around a time-constant expected value ($\dot{\phi}^{M^{a} \rightarrow O^{0}}_{o}$) with covariance matrix $\Sigma_{\phi^{M^{a} \rightarrow O^{0}}}$:

@@

The observation model components for the freshwater juvenile life stage were fitted to three primary data types: (_i_) abundance of juveniles passing rotary smolt traps (located in each tributary, \todo{Include reference to document describing methods of juvenile passage estimation}) in either the fall or spring, (_ii_) survival of juveniles from several PIT tagging events to LGR \todo{Include reference to document describing methods of juvenile survival estimation based on PIT tags}, and (_iii_) the mean length of sampled individuals during two of the PIT tagging events \todo{Include reference to document describing methods of juvenile mean length estimation}.
For all types, the information supplied to the model were externally compiled estimates (e.g., survival from Cormack-Jolly-Seber models) and the estimated standard error was supplied to the model as a measure of observation error variability.
A more fully integrated model would involve constructing the likelihood based on the frequency of PIT tag detection histories, however we thought unnecessary given the pre-compiled nature of currently available estimates and their uncertainty  \todo{Consider citing Staton et al. (2017) integrated vs. sequential SSSRA paper here}.

@@

We assumed first year natural-origin^[All $o=\mathrm{NOR}$ in eq. \@ref(eq:phi-O0-O1-random), omitted for brevity.] ocean survival was a multivariate logit-normal random variable around a time-constant expected value ($\dot{\phi}^{O^{0} \rightarrow O^{1}}_{o=\mathrm{NOR},j}$) with covariance matrix $\Sigma_{\phi^{O^{0} \rightarrow O^{1}}}$, but included a lag-1 autoregressive process as a means to account for ocean conditions that may affect survival in a temporally non-independent fashion\todo{Include reference to papers that talk about ocean survival varying cyclically; e.g., Mantua et al. paper on PDO}:

@@

We treated realized maturation rates (i.e., the proportion of fish from a given brood year alive and in the ocean at the beginning of a year that make the return migration that year) as multivariate logit-normal random variables around time-constant expected values ($\dot{\psi}^{O^{w}}_{o,j}$; $w$ is the number of winters spent in the ocean) with covariance matrix $\Sigma_{\psi^{O^{w}}}$ (assumed common across origins):

\begin{equation}
\begin{aligned}
\logit\left(\boldsymbol{\psi}^{O^{1}}_{y,o,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\psi}}^{O^{1}}_{o,1:n_j}\right), \Sigma_{\psi^{O^{1}}}\right] \\
\logit\left(\boldsymbol{\psi}^{O^{2}}_{y,o,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\psi}}^{O^{2}}_{o,1:n_j}\right), \Sigma_{\psi^{O^{2}}}\right]
\end{aligned}
(\#eq:psi-random)
\end{equation}

@@

The abundances of ocean juveniles ($O^{1}_{y,o,j}$, $O^{2}_{y,o,j}$, and $O^{3}_{y,o,j}$) are organized by brood year, however they will return in different years to spawn (and be observed) because they are of different ages.
We placed returning mature fish into the correct year and age of return to the mouth of the Columbia River before the upstream migration ($R^{b}_{y,k,o,j}$) based on the following rule.
Fish spawned in year $y$ will return as the $k^{\mathrm{th}}$ possible age of maturity in year $y + A_{\mathrm{min}} + k - 1$ (age-3 is the first possible total age of maturity, thus $A_{\mathrm{min}}$ = 3).
As an example, fish spawned in brood year $y$ = 2000 will return in 2003 as age-3 ($k$ = 1), in 2004 as age-4 ($k$ = 2), or in 2005 as age-5 ($k$ = 3) -- see Table \@ref(tab:year-table) for a visual of this example.
Thus, the abundance of fish spawned in brood year $y$ returning as age-3, age-4, and age-5, respectively, was:

\begin{equation}
\begin{aligned}
R^{b}_{y+A_{\mathrm{min}} + 1 - 1,1,o,j} &= O^{1}_{y,o,j} \cdot \psi^{O^{1}}_{y,o,j} \\
R^{b}_{y+A_{\mathrm{min}} + 2 - 1,2,o,j} &= O^{2}_{y,o,j} \cdot \psi^{O^{2}}_{y,o,j} \\
R^{b}_{y+A_{\mathrm{min}} + 3 - 1,3,o,j} &= O^{3}_{y,o,j} \cdot \psi^{O^{3}}_{y,o,j} \\
\end{aligned}
(\#eq:get-Rbs)
\end{equation}

@@

Prior to reaching BON, we subject returning adults to harvest mortality from fisheries (harvest rate denoted by $U_{y,k,o}$).\todo{Insert information about where harvest rate estimates came from.}
We compiled the best available estimates but did not fit to them explicitly as part of the observation model (as we do with many other data sources) -- instead we passed them to the model as non-stochastic quantities that were assumed to be known without error (used in eq. \@ref(eq:get-Ra), below).

@@

Once fish reach BON, however, their survival to LGR has been monitored (since 2000; $y=10$) via PIT tags and so we did fit to those data and estimate the survival rate between BON and LGR as multivariate logit-normal random variables around expected values ($\dot{\phi}^{R^{b} \rightarrow R^{a}}_{o}$) with covariance matrix $\Sigma_{\phi^{R^{b} \rightarrow R^{a}}}$.
In years without these PIT tag counts, however, the survival from BON to LGR is confounded with ocean survival, and so we assumed BON to LGR survival was equal to the expected value in these years for parameter identifiability purposes:

@@

\begin{equation}
R^{a}_{y,k,o,j} = R^{b}_{y,k,o,j} \cdot (1 - U_{y,k,o}) \cdot \phi^{R^{b} \rightarrow R^{a}}_{y,o} + \left(G_{y,o,j} \cdot p^{G}_{k,j}\right)
(\#eq:get-Ra)
\end{equation}

@@

Upon arrival to the natal tributary, populations with a weir and hatchery program (all populations except the Minam River, indexed by $j = \mathrm{MIN}$) have some number of adults removed each year for broodstock and as a means to minimize natural spawning of hatchery-origin fish.
We supplied these weir removals (denoted by $B_{y,k,o,j}$) to the model as known without error and set all $B_{y,k,o,j=\mathrm{MIN}} = 0$.
Similarly, fish are harvested in the natal tributaries prior to spawning, and we additionally supplied estimates of fish harvested ($H_{y,k,o,j}$).
Thus, the abundance of adults that would have a chance to spawn each year was:

\begin{equation}
S^{b}_{y,k,o,j} = \max\left(R^{a}_{y,k,o,j} - B_{y,k,o,j} - H_{y,k,o,j}, 1\right)
(\#eq:get-Sb)
\end{equation}
